Epistasis is widespread in the genetic control of transcription in humans

### 2013-05-29 docs/manuscript/nature_letter.tex
writing continues. around 400 words left...

\subsubsection{Replication}
It is intrinsically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than for independent additive effects. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$). There was substantial evidence for overlap of interaction effects in all three datasets. Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}). The congruence of the epistatic networks in both datasets is shown in Figure \ref{fig:fireworks}. We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and of these 18 were significant in both. The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}), lending strong evidence for the veracity of these signals. We also attempted replication in the CHDWB dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.

\subsubsection{Patterns of epistasis}

\subsubsection{Characterisation of genetic variation}

\subsubsection{Functional annotation}

\section{Notes}

Results

- Hundreds of significant pairwise interactions controlling gene expression replicate in two independent studies

- The genotype-phenotype maps for the most significant replications are consistent in all three datasets

- The proportion of the variance that is due to interactions is generally smaller than additive. There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold

- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent

- Functional annotation shows that proteins are not being changed at the codon level
- There is strong enrichment for enhancers
- There is tissue specific transcription regulation for cis SNPs - these are regulatory
- Trans SNPs are not found in tissue specific transcription regions.
- General enrichment for regulatory motif modifications
- Names of specific factors that are enriched.

- No explicit mechanism arises from looking at GP maps

- Chromosome interactions?

@@

\begin{figure}[H]
	\caption{\textbf{Replication of genotype-phenotype maps in 2 cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets. Each tile represents the expression level for that SNP-SNP genotype class. Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.}
	\label{fig:gpmaps}
\end{figure}

@@

The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \cite{pmid22563384}. DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland. Full details of genotyping procedures are given in Medland et al. \cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.

@@

\subsection{normalisation}

>>>>

\subsubsection{Replication}
Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$). There was substantial evidence for overlap of interaction effects in all three datasets. Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}). The congruence of the epistatic networks in both datasets is shown in Figure \ref{fig:fireworks}. We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both. The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}). We also attempted replication in the CHDWB dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.

\subsubsection{Functional annotation}
- Functional annotation shows that proteins are not being changed at the codon level
- There is strong enrichment for enhancers
- There is tissue specific transcription regulation for cis SNPs - these are regulatory
- Trans SNPs are not found in tissue specific transcription regions.
- General enrichment for regulatory motif modifications
- Names of specific factors that are enriched.
- Chromosome interactions?

\subsubsection{Patterns of epistasis}
- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent 
- No explicit mechanism arises from looking at GP maps
- Describe MBNL1

\subsubsection{Characterisation of genetic variation}
- The proportion of the variance that is due to interactions is generally smaller than additive. There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold

@@

\begin{figure}[H]
	\caption{\textbf{Replication of genotype-phenotype maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each tile represents the expression level for that two-locus genotype class. Phenotypes for each interaction within each cohort are scaled to be between 0 and 1. Phenotypes are for gene transcript levels, where the gene's name is labelled in the panel columns. Panel rows are labelled for the dataset. There is a general trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}

@@

The Brisbane Systems Genetics Study (BSGS) comprises 862 individuals of European descent from 274 independent families \cite{pmid22563384}. DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland. Full details of genotyping procedures are given in Medland et al. \cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.

@@

\subsection{Normalisation}
